\begin{thebibliography}{99}
\bibitem{insag1} International Atomic Energy Agency, \emph{Summary Report on the
Post-Accident Review Meeting on the Chernobyl Accident}, Safety Series
No.~75-INSAG-1, IAEA, Vienna (1986).
\bibitem{insag7} International Atomic Energy Agency, \emph{INSAG-7: The Chernobyl
Accident---Updating of INSAG-1}, Safety Series No.~75-INSAG-7, IAEA, Vienna
(1992).
\bibitem{wna-cher} World Nuclear Association, \emph{Chernobyl Accident 1986},
information library (accessed 2026).
\bibitem{wna-seq} World Nuclear Association, \emph{Sequence of Events---Chernobyl
Accident Appendix~1} (accessed 2026).
\bibitem{wna-rbmk} World Nuclear Association, \emph{RBMK Reactors}, appendix to
\emph{Nuclear Power Reactors} (accessed 2026).
\bibitem{lamarsh} J.~R.~Lamarsh and A.~J.~Baratta, \emph{Introduction to Nuclear
Engineering}, 3rd ed., Prentice Hall (2001).
\bibitem{duderstadt} J.~J.~Duderstadt and L.~J.~Hamilton, \emph{Nuclear Reactor
Analysis}, Wiley (1976).
\bibitem{henry} A.~F.~Henry, \emph{Nuclear-Reactor Analysis}, MIT Press (1975).
\bibitem{hetrick} D.~L.~Hetrick, \emph{Dynamics of Nuclear Reactors}, University of
Chicago Press (1971).
\bibitem{keepin} G.~R.~Keepin, \emph{Physics of Nuclear Kinetics}, Addison-Wesley
(1965).
\bibitem{stacey} W.~M.~Stacey, \emph{Nuclear Reactor Physics}, 2nd ed.,
Wiley-VCH (2007).
\bibitem{ans51} ANSI/ANS-5.1, \emph{Decay Heat Power in Light Water Reactors},
American Nuclear Society.
\bibitem{ebr2} H.~P.~Planchon et al., ``The experimental breeder reactor~II
inherent shutdown and heat removal tests,'' \emph{Nuclear Engineering and Design}
(1986).
\bibitem{pyrk} K.~D.~Huff et al., \emph{PyRK: Python for Reactor Kinetics},
\texttt{https://pyrk.github.io}.
\bibitem{wadefujita} D.~C.~Wade and E.~K.~Fujita, Trends versus reactor size of passive reactivity shutdown and control performance,'' \emph{Nuclear Science and Engineering} \textbf{103}, 182 (1989).
\bibitem{hairerwanner} E.~Hairer and G.~Wanner, \emph{Solving Ordinary Differential Equations II: Stiff and Differential-Algebraic Problems}, 2nd ed., Springer (1996).
\bibitem{molervanloan} C.~Moler and C.~Van~Loan, Nineteen dubious ways to compute the exponential of a matrix, twenty-five years later,'' \emph{SIAM Review} \textbf{45}, 3 (2003).
\bibitem{todreas} N.~E.~Todreas and M.~S.~Kazimi, \emph{Nuclear Systems I: Thermal Hydraulic Fundamentals}, 2nd ed., CRC Press (2012).
\bibitem{waltar} A.~E.~Waltar, D.~R.~Todd, and P.~V.~Tsvetkov (eds.), \emph{Fast Spectrum Reactors}, Springer (2012).
\end{thebibliography}
